Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-08-05 writing/DraftExposurePaper/DraftExposurePaper.Rmd
Work some on draft

Here are results for comparing our method to using ground-based monitors. First, we looked at several counties, comparing estimates of storm-associated rain using NLDAS-2 versus ground-based monitors (Figure [x]).

[Add in figures by county here for county averages of ground-based monitors versus our method. (for example, Miami-Dade figure)]

@@

# load("MWR_gauge_rain_data/alex_gauge.Rdata")
# alex_mwr <- plot_mwr_compare(alex_gauge)

# grid.arrange(allison_mwr, frances_mwr, ike_mwr, alex_mwr, ncol = 2)
grid.arrange(frances_wind_mwr, ike_wind_mwr,
             frances_wind_mwr, ike_wind_mwr, ncol = 2)
```

@@

We determined the average number of hits per year for each eastern US county for the exposure criteria based on distance (Figure \ref{fig:distance_hits}), rain (Figure \ref{fig:rain_hits}), and wind (Figure \ref{fig:wind_hits}), as well as the total number of hits based on storm-associated events listed in the NOAA Storm Events database for flood events (Figure \ref{fig:flood_hits}), wind events (Figure \ref{fig:wind_event_hits}), and tornado events (Figure \ref{fig:tornado_hits}) (all of these maps based on NOAA Storm Events are restricted to storms between 1996 and 2011).

@@

```{r wind_event_hits, echo = FALSE, fig.width = 8, fig.height = 6, fig.align = "center", fig.cap = "Average number of hits per year for eastern US counties based on the county having a storm-associated wind event listed in the NOAA Storm Events database, for storms from 1996 to 2011. Light red lines show the storm tracks over the same period. Red lines show the storm tracks for the top 5 storms in terms of number of counties classified as 'exposed' to the storm based on this criterion over the same period (Allison, 2001: 435 counties; Arlene, 2005: 360 counties; Danny, 1997: 310 counties; Cindy, 2005: 189 counties; Cristobal, 2008: 183 counties)."}
wind_event_exposure <- county_events(counties = my_fips,
                              start_year = 1996, end_year = 2011,
                              event_type = "wind")
map_total_hits(wind_event_exposure, my_fips = my_fips, 
               eastern_states = eastern_states,
               title = "Hits per county based on wind events")
```

@@

We also mapped geographic patterns in the number of hits per county for distance, rain, and wind exposure when we incorporated the intensity of exposure for each county-storm pair (Figures \ref{fip:top_dist_map}, \ref{fip:top_rain_map}, \ref{fip:top_wind_map}); Tables \ref{tab:top_distance}, \ref{tab:top_rain}, \ref{tab:top_wind}). For these maps, we only plotted "hits" for the top 2,000 county-storm pairs (i.e., the 2,000 storm-county pairs with smallest distance to storm's closest approach, in the case of distance-based exposure, or the 2,000 storm-county pairs with highest precipitation, in the case of rain-based exposure).

@@

```{r echo = FALSE, warning=FALSE, message=FALSE}
plot_top_x <- function(exposure_data, x = 2000, desc = TRUE,
                       exposure_var, eastern_states = eastern_states,
                       my_fips = my_fips, title){
        missing_fips <- my_fips[!(my_fips %in% exposure_data$fips)]
        missing_counties <- data.frame(fips = as.numeric(missing_fips),
                                       count = 0)

@@

        exposure_x <- exposure_data %>%
                arrange(desc(exp_var)) %>%
                slice(1:x) %>% 
                group_by(fips) %>%
                summarise(count = n()) %>%
                rbind(missing_counties) %>%
                rename(region = fips) %>%
                mutate(region = as.numeric(region),
                       count = as.numeric(count))

@@

```{r}
rain_exposure$county_storm <- transmute(rain_exposure,
                                        county_storm = paste0(fips, " ",
                                                              storm_id))
distance_exposure$county_storm <- transmute(distance_exposure,
                                            county_storm = paste0(fips, " ",
                                                                  storm_id))
all_exposure <- rain_exposure %>% 
        dplyr::select(county_storm, tot_precip) %>%
        dplyr::full_join(distance_exposure, by = "county_storm")  %>%
        dplyr::mutate(rain_exposure =  !is.na(tot_precip) & 
                              tot_precip >= 75,
                      dist_exposure = !is.na(storm_dist) & 
                              storm_dist < 100,
                      both_exposures = rain_exposure & dist_exposure,
                      rain_only = rain_exposure & !dist_exposure,
                      dist_only = !rain_exposure & dist_exposure,
                      neither_exposure = !rain_exposure & !dist_exposure) %>%
        dplyr::summarize(both_exposure = sum(both_exposures),
                         rain_only = sum(rain_only),
                         dist_only = sum(dist_only),
                         neither_exposure = sum(neither_exposure))
all_exposure

@@

One interesting contrast is between storms that were noteworthy for direct wind-related deaths (Anderew in 1992, Hugo in 1989) versus storms noteworthy for direct freshwater flooding-related deaths (Floyd in 1999, Alberto in 1994, Charley in 1998) for the 1970--1999 period [@Rappaport2000]. (Another study notes the following storms were notable for inland impacts-- Hugo, 1989 (wind and rain); Floyd, 1999 (flooding); Allison, 2001 (flooding); Fran, 1996 (wind) [@Kruk2010].) Here are two examples, one of each of these two types:

```{r fig.width = 6, fig.height = 8, echo = FALSE}
storm <- "Hugo-1989"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                          rain_limit = 75,
                                          dist_limit = 500)) + 
        ggtitle(paste0(storm, ", Rain-based exposure"))
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_wind_exposure(storm = storm, 
                                           wind_limit = 15)) + 
        ggtitle(paste0(storm, ", Wind-based exposure"))

```{r fig.width = 6, fig.height = 8, echo = FALSE}
storm <- "Allison-2001"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                          rain_limit = 75,
                                          dist_limit = 500)) + 
        ggtitle(paste0(storm, ", Rain-based exposure"))
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_wind_exposure(storm = storm, 
                                           wind_limit = 15)) + 
        ggtitle(paste0(storm, ", Wind-based exposure"))

@@

Here is a map of the rain totals for Frances:

```{r echo = FALSE}
map_tracks(storms = "Frances-2004", plot_points = FALSE, 
                plot_object = map_counties(storm = "Frances-2004", 
                                           metric = "rainfall")) + 
        ggtitle(paste("Frances-2004", "rainfall"))
```

Here is a table of the top 10 storms by number of counties exposed, based on a wind metric of 18 m/s or greater (e.g., tropical storm strength or greater):

```{r echo = FALSE, message = FALSE, warning = FALSE}
all_wind <- county_wind(counties = my_fips, start_year = 1988, 
                        end_year = 2011, wind_limit = 18)

@@

Here is a table of the top 10 storms by number of counties exposed, based on a wind metric of 33 m/s or greater (e.g., Hurricane strength, category 1 or greater):

```{r echo = FALSE, message = FALSE, warning = FALSE}
all_wind <- county_wind(counties = my_fips, start_year = 1988, 
                        end_year = 2011, wind_limit = 33)

hu_exposed <- all_wind %>%
        group_by(storm_id) %>%
        summarize(number_of_counties_exposed = n()) %>%
        arrange(desc(number_of_counties_exposed)) %>%
        top_n(10)

hu_exposed %>%
        mutate(storm_id = gsub("-", " (", storm_id),
               storm_id = paste0(storm_id, ")")) %>%
        kable(col.names = c("Storm", "# of exposed counties"),
              caption = "Top 10 storms based on the number of counties exposed, using a metric of wind.")
```

Here is a table of the top 10 storms by number of counties exposed, based on a wind metric of 50 m/s or greater (e.g., Intense Hurricane strength, category 3 or greater):

```{r echo = FALSE, message = FALSE, warning = FALSE}
all_wind <- county_wind(counties = my_fips, start_year = 1988, 
                        end_year = 2011, wind_limit = 50)

cat3_exposed <- all_wind %>%
        group_by(storm_id) %>%
        summarize(number_of_counties_exposed = n()) %>%
        arrange(desc(number_of_counties_exposed)) %>%
        top_n(10)

cat3_exposed %>%
        mutate(storm_id = gsub("-", " (", storm_id),
               storm_id = paste0(storm_id, ")")) %>%
        kable(col.names = c("Storm", "# of exposed counties"),
              caption = "Top 10 storms based on the number of counties exposed, using a metric of wind.")
```

For all of the above exposure categories, Lili (1996), Karl (2004), and Ophelia (2005) remained the top 3 storms by number of counties exposed based on the various wind exposure metrics, each with 2396 counties exposed [JMF-- I think something weird is going on here].

@@

For storms in 1996 or later, when flooding reports are included in NOAA's Storm Events data, exposure patterns based on floods reports generally pick up similar regions as rain-based exposure, and in some cases (e.g., Floyd, 1999), counties classified as exposed using rain-based exposure are very similar to the counties in which flood events were recorded. For example, here are comparisons for some of the worst storms in terms of rain-based exposures:

```{r echo = FALSE, fig.width = 6, fig.height = 8}
storm <- "Frances-2004"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Frances, 2004")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_event_exposure(storm = storm, 
                                           event_type = "flood")) + 
        ggtitle("Flood events during Frances, 2004")
grid.arrange(a, b, ncol = 1)
```

```{r echo = FALSE, fig.width = 6, fig.height = 8}
storm <- "Floyd-1999"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Floyd, 1999")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_event_exposure(storm = storm, 
                                           event_type = "flood")) + 
        ggtitle("Flood events during Floyd, 1999")
grid.arrange(a, b, ncol = 1)
```

>>>>

Here are results for comparing our method to using ground-based monitors. First, we looked at several counties, comparing estimates of storm-associated rain using NLDAS-2 versus ground-based monitors (Figure \ref{fig:compare_rain_ex_counties}).

```{r compare_rain_ex_counties, echo = FALSE, fig.width = 10, fig.height = 10, efig.align = "center", fig.cap = "Comparison of rain estimates for example counties."}
plot_county_rain_compare <- function(ex_fips, ex_dir, ex_title, 
                                     get_data = FALSE){
        if(get_data){
                county_timeseries(ex_fips, percent_coverage = 0,
                                  date_min = "1988-01-01",
                                  date_max = "2011-12-31",
                                  var = "PRCP",
                                  out_directory = ex_dir)
        }

        check_dates <- closest_dist %>%
                dplyr::filter(fips == ex_fips) %>%
                dplyr::select(-storm_dist) %>%
                dplyr::mutate(closest_date = ymd(closest_date)) %>%
                dplyr::rename(day_0 = closest_date) %>%
                dplyr::mutate(fips = as.integer(fips),
                              day_0 = day_0 + days(0),
                              day_b1 = day_0 - days(1),
                              day_b2 = day_0 - days(2),
                              day_a1 = day_0 + days(1)) %>%
                dplyr::select(storm_id, day_b2, day_b1, day_0, day_a1) %>%
                tidyr::gather(key = lag, value = day, -storm_id) %>%
                dplyr::rename(date = day)
        all_dates <- unique(check_dates$date)

        ex_weather <- readRDS(paste0(ex_dir, ex_fips, ".rds"))
        ex_weather <- as.data.frame(ex_weather$averaged)

        ex_weather <- ex_weather %>%
                dplyr::filter(date %in% all_dates) %>%
                dplyr::right_join(check_dates, by = "date") %>%
                dplyr::group_by(storm_id) %>%
                dplyr::summarize(prcp = sum(prcp),
                                ave_n = mean(prcp_reporting))

        ex_rain <- county_rain(counties = ex_fips,
                               start_year = 1988, end_year = 2011,
                               rain_limit = 0, dist_limit = 2000) %>%
                full_join(ex_weather, by = "storm_id") %>%
                filter(!is.na(prcp) & !is.na(tot_precip))

        plot_range <- range(ex_rain[ , c("prcp", "tot_precip")],
                            na.rm = TRUE)

        ex_plot <- ggplot(ex_rain, aes(x = tot_precip, y = prcp)) +
                geom_hline(aes(yintercept = 75), color = "lightgray") +
                geom_vline(aes(xintercept = 75), color = "lightgray") +
                geom_abline(aes(intercept = 0, slope = 1), color = "gray",
                    alpha = 0.5) +
                geom_point(alpha = 0.5) +
                theme_few() +
                scale_size_continuous(guide = "none") +
                xlab("Rainfall (mm)  from\nNLDAS-2 data") +
                ylab("Rainfall (mm)\nfrom monitors") +
                xlim(plot_range) + ylim(plot_range) + 
                ggtitle(ex_title)
        return(ex_plot)
}

a <- plot_county_rain_compare(ex_fips = "12086", ex_dir = "dade_data/",
                         ex_title = "Miami-Dade, FL")

b <- plot_county_rain_compare(ex_fips = "22071", ex_dir = "orleans_data/",
                         ex_title = "Orleans Parish, LA")

c <- plot_county_rain_compare(ex_fips = "45019", ex_dir = "charleston_data/",
                         ex_title = "Charleston County, SC")

d <- plot_county_rain_compare(ex_fips = "37183", ex_dir = "wake_data/",
                         ex_title = "Wake County, NC")

e <- plot_county_rain_compare(ex_fips = "13121", ex_dir = "fulton_data/",
                         ex_title = "Fulton County, GA")

f <- plot_county_rain_compare(ex_fips = "24005", ex_dir = "baltimore_data/",
                         ex_title = "Baltimore County, MD")

g <- plot_county_rain_compare(ex_fips = "42101",
                              ex_dir = "philadelphia_data/",
                         ex_title = "Philadelphia County, PA")

h <- plot_county_rain_compare(ex_fips = "48201", ex_dir = "harris_data/",
                         ex_title = "Harris County, TX")

i <- plot_county_rain_compare(ex_fips = "01097", ex_dir = "mobile_data/",
                         ex_title = "Mobile County, AL", get_data = TRUE)

grid.arrange(a, b, c, d, e, f, g, h, i, ncol = 3)
```

@@

load("MWR_gauge_wind_data/alex_wind.Rdata")
alex_wind <- alex_wind %>%
        filter(!is.na(max_sust)) %>%
        mutate(wind_sust = convert_wind_speed(max_sust, "knots", "mps")) %>%
        select(storm_id, fips, wind_sust)
alex_wind_mwr <- plot_mwr_compare_wind(alex_wind)

grid.arrange(frances_wind_mwr, ike_wind_mwr,
             frances_wind_mwr, alex_wind_mwr, ncol = 2)
```

@@

We determined the average number of hits per year for each eastern US county for the exposure criteria based on distance (Figure \ref{fig:distance_hits}), rain (Figure \ref{fig:rain_hits}), and wind (Figure \ref{fig:wind_hits}), as well as the total number of hits based on storm-associated events listed in the NOAA Storm Events database for flood events (Figure \ref{fig:flood_hits}) and tornado events (Figure \ref{fig:tornado_hits}) (all of these maps based on NOAA Storm Events are restricted to storms between 1996 and 2011).

@@

@@

Some storms can very substantially in their exposure patterns when different metrics are used. For example, Tropical Storm Allison (2001) was notable for its rainfall rather than wind intensity; as a result, many more counties are classified as "exposed" to that storm when the exposure classification is based on rainfall rather than wind (Figure \ref{fig:allison_wind_rain_exp}). Conversely, Hurricane Hugo (1989) was more notable for wind than rainfall, which is reflected in more counties assessed as "exposed" based on wind than rain (Figure \ref{fig:hugo_wind_rain_exp}).

[One interesting contrast is between storms that were noteworthy for direct wind-related deaths (Anderew in 1992, Hugo in 1989) versus storms noteworthy for direct freshwater flooding-related deaths (Floyd in 1999, Alberto in 1994, Charley in 1998) for the 1970--1999 period [@Rappaport2000]. (Another study notes the following storms were notable for inland impacts-- Hugo, 1989 (wind and rain); Floyd, 1999 (flooding); Allison, 2001 (flooding); Fran, 1996 (wind) [@Kruk2010].) Here are two examples, one of each of these two types:]

```{r allison_wind_rain_exp, fig.width = 6, fig.height = 8, echo = FALSE, fig.align = "center", fig.cap = "Counties that were classified as 'exposed' to Tropical Storm Allison (2001) based on rain (top) and wind (bottom) exposure criteria. Rain-based exposure was assessed based on storm-associated rainfall of 75 millimeters or more in the county and a distance of 500 kilometers or less between the county's population mean center and the storm's closest approach. Wind-based exposure was based on maximum sustained winds of 15 meters per second or more in the county during the storm."}
storm <- "Allison-2001"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                          rain_limit = 75,
                                          dist_limit = 500)) + 
        ggtitle(paste0(storm, ", Rain-based exposure"))
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_wind_exposure(storm = storm, 
                                           wind_limit = 15)) + 
        ggtitle(paste0(storm, ", Wind-based exposure"))

```{r hugo_wind_rain_exp, fig.width = 6, fig.height = 8, echo = FALSE, fig.align = "center", fig.cap = "Counties that were classified as 'exposed' to Hurricane Hugo (1989) based on rain (top) and wind (bottom) exposure criteria. Rain-based exposure was assessed based on storm-associated rainfall of 75 millimeters or more in the county and a distance of 500 kilometers or less between the county's population mean center and the storm's closest approach. Wind-based exposure was based on maximum sustained winds of 15 meters per second or more in the county during the storm."}
storm <- "Hugo-1989"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                          rain_limit = 75,
                                          dist_limit = 500)) + 
        ggtitle(paste0(storm, ", Rain-based exposure"))
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_wind_exposure(storm = storm, 
                                           wind_limit = 15)) + 
        ggtitle(paste0(storm, ", Wind-based exposure"))

As another example, exposure classifications based on flood events in some cases align well with rain-based classifications (e.g., for Hurricane Floyd in 1999; Figure \ref{fig:floyd_rain_flood_exp}) but for other storms only a small percent of the counties assessed as "exposed" based on rain criteria are also assessed as "exposed" based on flood exposure (e.g., Hurricane Frances, 2004; Figure \ref{fig:frances_rain_flood_exp}).

[For storms in 1996 or later, when flooding reports are included in NOAA's Storm Events data, exposure patterns based on floods reports generally pick up similar regions as rain-based exposure, and in some cases (e.g., Floyd, 1999), counties classified as exposed using rain-based exposure are very similar to the counties in which flood events were recorded. For example, here are comparisons for some of the worst storms in terms of rain-based exposures:]

```{r frances_rain_flood_exp, echo = FALSE, fig.width = 6, fig.height = 8, fig.align = "center", fig.cap = "Counties that were classified as 'exposed' to Hurricane Frances (2004) based on rain (top) and flood event (bottom) exposure criteria. Rain-based exposure was assessed based on storm-associated rainfall of 75 millimeters or more in the county and a distance of 500 kilometers or less between the county's population mean center and the storm's closest approach. Flood event-based exposure was based on the county have a flood event listing in the NOAA Storm Events database and being within [x] kilometers of the storm's track."}
storm <- "Frances-2004"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Frances, 2004")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_event_exposure(storm = storm, 
                                           event_type = "flood")) + 
        ggtitle("Flood events during Frances, 2004")
grid.arrange(a, b, ncol = 1)
```

```{r floyd_rain_flood_exp, echo = FALSE, fig.width = 6, fig.height = 8, fig.align = "center", fig.cap = "Counties that were classified as 'exposed' to Hurricane Floyd (1999) based on rain (top) and flood event (bottom) exposure criteria. Rain-based exposure was assessed based on storm-associated rainfall of 75 millimeters or more in the county and a distance of 500 kilometers or less between the county's population mean center and the storm's closest approach. Flood event-based exposure was based on the county have a flood event listing in the NOAA Storm Events database and being within [x] kilometers of the storm's track."}
storm <- "Floyd-1999"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Floyd, 1999")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_event_exposure(storm = storm, 
                                           event_type = "flood")) + 
        ggtitle("Flood events during Floyd, 1999")
grid.arrange(a, b, ncol = 1)
```

We mapped geographic patterns in the number of hits per county for distance, rain, and wind exposure when we incorporated the intensity of exposure for each county-storm pair (Figures \ref{fig:top_dist_map}, \ref{fig:top_rain_map}, \ref{fig:top_wind_map}); Tables \ref{tab:top_distance}, \ref{tab:top_rain}, \ref{tab:top_wind}). For these maps, we only plotted "hits" for the top 2,000 county-storm pairs (i.e., the 2,000 storm-county pairs with smallest distance to storm's closest approach, in the case of distance-based exposure, or the 2,000 storm-county pairs with highest precipitation, in the case of rain-based exposure).

@@

```{r echo = FALSE, warning=FALSE, message=FALSE}
plot_top_x <- function(exposure_data, x = 2000, desc = TRUE,
                       exposure_var, eastern_states = eastern_states,
                       my_fips = my_fips, title){

@@

        exposure_x <- exposure_data %>%
                arrange(desc(exp_var)) %>%
                slice(1:x) %>% 
                group_by(fips) %>%
                summarise(count = n())

        zero_fips <- my_fips[!(my_fips %in% exposure_x$fips)]
        zero_counties <- data.frame(fips = as.numeric(zero_fips),
                                       count = 0)

        exposure_x <- exposure_x %>%
                rbind(zero_counties) %>%
                rename(region = fips) %>%
                mutate(region = as.numeric(region),
                       count = as.numeric(count))

@@

```{r echo = FALSE}
rain_exposure$county_storm <- transmute(rain_exposure,
                                        county_storm = paste0(fips, " ",
                                                              storm_id))
distance_exposure$county_storm <- transmute(distance_exposure,
                                            county_storm = paste0(fips, " ",
                                                                  storm_id))
all_exposure <- rain_exposure %>% 
        dplyr::select(county_storm, tot_precip) %>%
        dplyr::full_join(distance_exposure, by = "county_storm")  %>%
        dplyr::mutate(rain_exposure =  !is.na(tot_precip) & 
                              tot_precip >= 75,
                      dist_exposure = !is.na(storm_dist) & 
                              storm_dist < 100,
                      both_exposures = rain_exposure & dist_exposure,
                      rain_only = rain_exposure & !dist_exposure,
                      dist_only = !rain_exposure & dist_exposure,
                      neither_exposure = !rain_exposure & !dist_exposure) %>%
        dplyr::summarize(both_exposure = sum(both_exposures),
                         rain_only = sum(rain_only),
                         dist_only = sum(dist_only),
                         neither_exposure = sum(neither_exposure))
all_exposure

@@

@@

Here is a table of the top 10 storms by number of counties exposed, based on maximum sustained winds of 15 m / s or greater (e.g., tropical storm strength or greater):

```{r echo = FALSE, message = FALSE, warning = FALSE}
all_wind <- county_wind(counties = my_fips, start_year = 1988, 
                        end_year = 2011, wind_limit = 15)

@@

@@
